A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-02 introProtocol.tex
Finished first part of Reviewer 1's comments. Still a work in progress

	\textcolor{red}{We should also incorporate CSMA/CA$_{\text{FS}}$??.}

@@

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

@@

	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol. \textcolor{red}{Is this still true?}

@@

	\textcolor{red}{\subsection{Real hardware implementation} 
	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}

	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

>>>>

@@

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$. \textcolor{red}{This figure should also show CSMA/CA$_{\text{FS}}$}.}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

@@

	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.

@@

%	\textcolor{red}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%	
	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

### 2015-10-02 journal.tex
Finished first part of Reviewer 1's comments. Still a work in progress

%A more simple output, useful when involving people from different affiliations
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}}%\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
 % }

@@

\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
L. Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J. Barcelo (\href{mfullyailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B. Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})

@@

The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.

\textcolor{red}{In this part we should also assess the channel error scenario and the real implementation with OpenFWWF.}

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\textcolor{red}{\item The effect of channel errors over collision-free schedules and how to solve the issue of converging rapidly to the biggest cycle length with the Schedule Reset mechanism.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item Implementation of CSMA/ECA in real hardware.
\end{itemize}

@@

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis. Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

>>>>

%A more simple output, useful when involving people from different affiliations
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}, Francesco Gringoli\IEEEauthorrefmark{0}}}%\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
 % }

@@

\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
L. Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J. Barcelo (\href{mfullyailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B. Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.\\ F. Gringoli (\href{mailto:francesco.gringoli@unibs.it}{francesco.gringoli@unibs.it}) is with the Dipartamento di Elettronica per l’Automazione, Universit\`{a} degli Studi di Brescia, Brescia, Italy.}
%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})

@@

The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of channel errors and imperfect clocks over the deterministic backoff mechanism is also lacking.

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\textcolor{red}{\item The effect of channel errors over collision-free schedules and how to solve the issue of converging rapidly to the biggest cycle length with the Schedule Reset mechanism.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

@@

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

### 2015-10-02 realImplementation.tex
Finished first part of Reviewer 1's comments. Still a work in progress

\section{Real hardware implementation}
The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).

>>>>

\section{Real hardware implementation}\label{EDCA}

\textcolor{red}{The new section should assume that the implementation has already been done. Further, it should overview:}

\textcolor{red}{\begin{itemize}
	\item Why do we need it?
	\item Are there other alternatives?
	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
\end{itemize}}

\textcolor{red}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}

The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).

### 2015-10-02 simulations.tex
Finished first part of Reviewer 1's comments. Still a work in progress

\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.

@@

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item There are no external interference or channel errors. \textcolor{red}{Remove}.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: \textcolor{red}{channel imperfections}. Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot. \textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}

	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures also show the standard deviation.

@@

	\textcolor{red}{\subsection{Channel errors} Here we explain how we implement our channel errors model}.

>>>>

\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. \textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions}. Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.

@@

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behaviour.
	%Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot. \textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}
	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures also show the standard deviation.

@@

	\textcolor{red}{\subsection{Channel errors} 
	We model channel errors by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each packet individually and independently. Therefore, a A-MPDU transmission will be considered a failure is all packets in the aggregation are affected by the channel error probability.}
